Equalizer Programming Guide

### 2007-08-31 Developer/ProgrammingGuide/paper.tex
New features: - divide Config::init into startInit and finishInit - more functionality in monitor primitive
Cleanups:
- replace error-prone _pendingRequestID synchronization during config
  init and exit with state monitoring

Bugfixes:
- Don't use alloca during big package send
- Correctly report config exit results

Documentation:
- Update Programming Guide
- Update data transmission API

[ ] May break build
[ ] Breaks existing applications
[x] Bugfix
[x] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

\usepackage{tabularx,graphicx,a4,listings}

@@

\title{Equalizer Programming Guide}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.3, \today
}

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.3         & Aug 26, 2007 & application and render client code described\\
    0.2         & Aug 20, 2007 & main function described\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline \multicolumn{3}{c}{\small
      \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
      {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}\\
  \end{tabularx}}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\columnwidth]{cave.pdf}{\caption{\label{fConfig}An
      example configuration}}
\end{figure}

@@

{\small\begin{lstlisting}
int main( int argc, char** argv )
{
    // 1. parse arguments
    eqPly::LocalInitData initData;
    initData.parseArguments( argc, argv );
\end{lstlisting}}

@@

{\small\begin{lstlisting}
    // 2. Equalizer initialization
    NodeFactory nodeFactory;
    if( !eq::init( argc, argv, &nodeFactory ))
    {
        EQERROR << "Equalizer init failed" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

{\small\begin{lstlisting}
class NodeFactory : public eq::NodeFactory
{
public:
    virtual eq::Config*  createConfig()  { return new eqPly::Config; }
    virtual eq::Node*    createNode()    { return new eqPly::Node; }
    virtual eq::Pipe*    createPipe()    { return new eqPly::Pipe; }
    virtual eq::Window*  createWindow()  { return new eqPly::Window; }
    virtual eq::Channel* createChannel() { return new eqPly::Channel; }
};
\end{lstlisting}}

@@

{\small\begin{lstlisting}
    // 3. initialization of local client node
    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
    if( !client->initLocal( argc, argv ))
    {
        EQERROR << "Can't init client" << endl;
        eq::exit();
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

{\small\begin{lstlisting}
    // 4. run client
    const int ret = client->run();
\end{lstlisting}}

@@

{\small\begin{lstlisting}
    // 5. cleanup and exit
    client->exitLocal();
    client = 0;

@@

{\small\begin{lstlisting}
int Application::run()
{
    // 1. connect to server
    RefPtr<eq::Server> server = new eq::Server;
    if( !connectServer( server ))
    {
        EQERROR << "Can't open server" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

{\small\begin{lstlisting}
    // 2. choose config
    eq::ConfigParams configParams;
    Config* config = static_cast<Config*>(server->chooseConfig( configParams ));

@@

{\small\begin{lstlisting}
    // 3. init config
    eqBase::Clock clock;

@@

When the configuration was successfully initialized, the actual main
loop is executed. The main loop runs until the user exits the
configuration or a maximum number of frames, specified as a command-line
argument, has been rendered. The latter is useful for benchmarks. The
\textsf{Clock} is reused for measuring the overall performance. A new
frame is started using \textsf{Config::startFrame} and a frame is
finished using \textsf{Config::endFrame}.

@@

The frame finished is \textsf{latency} frames behind the frame
started. The latency is specified in the configuration file, and allows
several outstanding frames whose tasks are already queued in the node
and pipe threads for execution. This allows overlapped execution and
minimizes idle times.

When the main loop is finished, \textsf{Config::finishAllFrames} catches
up with the latency. It returns after all outstanding frames have been
rendered:

{\small\begin{lstlisting}
    // 4. run main loop
    uint32_t maxFrames = _initData.getMaxFrames();

@@

{\small\begin{lstlisting}
    // 5. exit config
    clock.reset();
    config->exit();
    EQLOG( eq::LOG_CUSTOM ) << "Exit took " << clock.getTimef() << " ms" <<endl;

@@

In the second and third case, when the executable is used as a render
client, \textsf{Client::initLocal} never returns. In order to keep the
client resident, the \textsf{eqPly} example overrides the client loop to
keep it running beyond one configuration run:

{\small\begin{lstlisting}
bool Application::clientLoop()
{
    if( !_initData.isResident( )) // execute only one config run
        return eq::Client::clientLoop();

@@

\if 0
{\small\begin{lstlisting}
\end{lstlisting}}
\fi

>>>>

\usepackage{tabularx,graphicx,a4,listings,wrapfig,subfigure}

@@

\title{Equalizer Programming Guide}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.4, \today
}

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.4         & Aug 31, 2007 & added distributed objects\\
    0.3         & Aug 26, 2007 & added application and render client\\
    0.2         & Aug 20, 2007 & added main function\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline \multicolumn{3}{c}{\small
      \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
      {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}\\
  \end{tabularx}}

@@

\begin{wrapfigure}{r}{.5\textwidth}
  \includegraphics[width=.5\textwidth]{cave.pdf}{\caption{\label{fConfig}An
      example configuration}}
\end{wrapfigure}

@@

{\footnotesize\begin{lstlisting}
int main( int argc, char** argv )
{
    // 1. parse arguments
    eqPly::LocalInitData initData;
    initData.parseArguments( argc, argv );
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 2. Equalizer initialization
    NodeFactory nodeFactory;
    if( !eq::init( argc, argv, &nodeFactory ))
    {
        EQERROR << "Equalizer init failed" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
class NodeFactory : public eq::NodeFactory
{
public:
    virtual eq::Config*  createConfig()  { return new eqPly::Config; }
    virtual eq::Node*    createNode()    { return new eqPly::Node; }
    virtual eq::Pipe*    createPipe()    { return new eqPly::Pipe; }
    virtual eq::Window*  createWindow()  { return new eqPly::Window; }
    virtual eq::Channel* createChannel() { return new eqPly::Channel; }
};
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 3. initialization of local client node
    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
    if( !client->initLocal( argc, argv ))
    {
        EQERROR << "Can't init client" << endl;
        eq::exit();
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 4. run client
    const int ret = client->run();
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 5. cleanup and exit
    client->exitLocal();
    client = 0;

@@

{\footnotesize\begin{lstlisting}
int Application::run()
{
    // 1. connect to server
    RefPtr<eq::Server> server = new eq::Server;
    if( !connectServer( server ))
    {
        EQERROR << "Can't open server" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 2. choose config
    eq::ConfigParams configParams;
    Config* config = static_cast<Config*>(server->chooseConfig( configParams ));

@@

{\footnotesize\begin{lstlisting}
    // 3. init config
    eqBase::Clock clock;

@@

When the configuration was successfully initialized, the actual main
loop is executed. The main loop runs until the user exits the
configuration or a maximum number of frames, specified as a command-line
argument, has been rendered. The latter is useful for benchmarks. The
\textsf{Clock} is reused for measuring the overall performance. A new
frame is started using \textsf{Config::startFrame} and a frame is
finished using \textsf{Config::finishFrame}.

@@

\begin{wrapfigure}{r}{.5\textwidth}
  \subfigure[]{\label{fSync}\includegraphics[width=.22\textwidth]{sync}}\hfil
  \subfigure[]{\label{fAsync}\includegraphics[width=.22\textwidth]{async}}%
  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames for which the
tasks are already queued in the node and pipe threads for
execution. This enables overlapped execution and minimizes idle
times. The first \textsf{latency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. Figure~\ref{fSyncAsync} shows the execution of (hypothetical)
rendering tasks without latency~(\ref{fSync}) and with a latency of
one~(\ref{fAsync}).

When the main loop is finished, \textsf{Config::finishAllFrames} catches
up with the latency. It returns after all outstanding frames have been
rendered, and is needed here to provide an accurate measurement of the
framerate:

{\footnotesize\begin{lstlisting}
    // 4. run main loop
    uint32_t maxFrames = _initData.getMaxFrames();

@@

{\footnotesize\begin{lstlisting}
    // 5. exit config
    clock.reset();
    config->exit();
    EQLOG( eq::LOG_CUSTOM ) << "Exit took " << clock.getTimef() << " ms" <<endl;

@@

In the second and third case, when the executable is used as a render
client, \textsf{Client::initLocal} never returns. Therefore the
application's main loop is never executed. In order to keep the client
resident, the \textsf{eqPly} example overrides the client loop to keep
it running beyond one configuration run:

{\footnotesize\begin{lstlisting}
bool Application::clientLoop()
{
    if( !_initData.isResident( )) // execute only one config run
        return eq::Client::clientLoop();

@@

\subsection{Distributed Objects}

Equalizer provides distributed objects which help implementing data
distribution in a graphics cluster. The master version of a distributed
object is registered with a \textsf{eqNet::Session}, which assigns a
session-unique identifier to the object. Other nodes can map their
instance of the object to this identifier, thus synchronizing the
object's data with the remotely registered object.

Distributed object are created by subclassing from
\textsf{eqNet::Object}. Distributed objects can be static (immutable) or
dynamic. Dynamic objects are versioned.

The \textsf{eqPly} example has a static distributed object to provide
initial data to all rendering nodes, as well as a versioned object to
provide frame-specific data, such as the camera position, to the
rendering methods.

\subsubsection{InitData - a Static Distributed Object}

The \textsf{InitData} class holds a couple of parameters needed during
initialization. These parameters never change during one configuration
run, and are therefore static.

A static distributed object either has to provide a pointer and size to
its data using \textsf{setInstanceData} or it has to implement
\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
approach can be used if all distributed member variables can be outlayed
in one contiguous block of memory. The second approach is used
otherwise.

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and deserializing its data. The serialization is done in
\textsf{getInstanceData}. The serialized data is cached using a
(non-distributed) member variable, which is cleared by each setter of
distributed data. Serializing the data is a simple matter of allocating
a large enough memory buffer, and copying the data into the buffer:

{\footnotesize\begin{lstlisting}
const void* InitData::getInstanceData( uint64_t* size )
{
    *size = sizeof( uint32_t ) + sizeof( eq::WindowSystem ) +
            _filename.length() + 1;
    if( _instanceData )
        return _instanceData;

    _instanceData = new char[ *size ];

    reinterpret_cast< uint32_t* >( _instanceData )[0] =  _frameDataID;
    reinterpret_cast< uint32_t* >( _instanceData )[1] =  _windowSystem;

    const char* string = _filename.c_str();
    memcpy( _instanceData + sizeof( uint64_t ), string, _filename.length()+1 );

    return _instanceData;
}
\end{lstlisting}}

After the data is no longer needed, \textsf{releaseInstanceData} is
called by Equalizer to allow the object to free the data. Since the data
is cached for further usage, the release method is not overwritten.

The memory buffer returned by \textsf{getInstanceData} is transferred to
the remote node during object mapping and passed to
\textsf{applyInstanceData} for deserialization. The deserialization
routine reads the data back into its own member variables:

{\footnotesize\begin{lstlisting}
void InitData::applyInstanceData( const void* data, const uint64_t size )
{
    EQASSERT( size > ( sizeof( _frameDataID ) + sizeof( _windowSystem )));

    _frameDataID  = reinterpret_cast< const uint32_t* >( data )[0];
    _windowSystem = reinterpret_cast< const eq::WindowSystem* >( data )[1];
    _filename     = static_cast<const char*>( data ) + sizeof( uint64_t );

    EQASSERT( _frameDataID != EQ_ID_INVALID );

    EQINFO << "New InitData instance" << endl;
}
\end{lstlisting}}

\subsubsection{FrameData - a Versioned Distributed Object}

Versioned objects have to override \textsf{isStatic} to return false, in
order for Equalizer to know that they should be versioned. Right now,
only the master instance of the object is writable, that is,
\textsf{eqNet::Object::commit} can be called to generate a new
version.

Upon \textsf{commit} the delta data from the previous version is sent to
all mapped slave instances. The data is queued on the remote node, and
is applied when the application calls \textsf{sync} to synchronize the
object to a new version. The \textsf{sync} method might block if a
version has not been committed or is still in transmission.

In addition to the instance data (de)serialization methods needed to map
an object, versioned objects may implement \textsf{pack} and
\textsf{unpack} to serialize or deserialize the changes since the last
version.

If the delta data happens to be layed out contiguously in memory,
\textsf{setDeltaData} might be used. The default implementation of
\textsf{pack} and \textsf{unpack} (de)serialize the delta data or the
instance data if no delta data has been specified.

The frame data is layed out in one anonymous structure in
memory. It also does not track changes since it is relatively small in
size and changes frequently. Therefore, for the instance and delta
date are the same and set in the constructor:

{\footnotesize\begin{lstlisting}
        FrameData()
            {
                reset();
                setInstanceData( &data, sizeof( Data ));
                EQINFO << "New FrameData " << std::endl;
            }
\end{lstlisting}}

\if 0
\subsection{Config}

\subsubsection{Config Initialization}

The \textsf{eqPly::Config} class holds
{\footnotesize\begin{lstlisting}
bool Config::init()
{
    // init distributed objects
    _frameData.data.color = _initData.useColor();
    registerObject( &_frameData );
    _initData.setFrameDataID( _frameData.getID( ));

    registerObject( &_initData );
\end{lstlisting}}
\fi

\if 0
{\footnotesize\begin{lstlisting}
\end{lstlisting}}
\fi
